Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-12-04 AuthorResponseToReviewers.Rmd

*The most crucial point is in my opinion the missing discussion of the influence of different data distributions. In the visualization community exists research on different descriptors for two-dimensional data called Scagnostics. When I started reading this article, I expected a thorough discussion of different data distributions/properties and the influence on different binning strategies and error outcomes.*

- We are very grateful that the review brought the literature on scagnostics to our attention; the work and scagnostic metrics formalized in the works by Wilkerson, Anand, Grosmann and Wills will be valuable to the discussion of binned scatterplots. Specifically scagnostic literature contributes to Section 5 of our paper in the exploration of loss under a variety of simulated distribution. You will find that a discussion of the relevance of scagnostics has been added immediately preceding the start of subsection 5.1. It is interesting that the striation score is so small, given that the data resolution issue can lead to artificially induced stripping is seen in so prominently in Figure 6.

*In Section 4.2, a discussion on contrast effects hindering the perception of the encoded values is missing. Nearby colored points, will change the color impression. These effects can be modelled and estimated and maybe interesting to investigate a little further.*

- Citations to refer readers to Bartelson & Breneman, Adelson, and Fairchild as example of work on contrast effects has been added to Section 4.2 and related future work has been added to the discussion in Section 6.

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Add also Bartelson and Brenemon to shade contrast list.

% It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates. The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from over-plotting, which can be immense for large data sets. When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location. This implicit loss of frequency information is then exacerbated when points are rendered as a circle with a radius large enough to also partially cover nearby points. In binned scatterplots, we have traded the exact locations for exact frequencies within bins. 
% 
Bin counts can be displayed using a continuous shading scale in HSV color space \citep{colorperception} and thus we can theoretically map frequency to shade perfectly. While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise. Color perception of shade is \ktm{influenced by the contrasting effect of surrounding shades}, allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend \ktm{\citep{adelson1993perceptual,fairchild2013color}}. See Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors. It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.  %Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.

>>>>

% It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates. The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from over-plotting, which can be immense for large data sets. When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location. This implicit loss of frequency information is then exacerbated when points are rendered as a circle with a radius large enough to also partially cover nearby points. In binned scatterplots, we have traded the exact locations for exact frequencies within bins. 
% 
Bin counts can be displayed using a continuous shading scale in HSV color space \citep{colorperception} and thus we can theoretically map frequency to shade perfectly. While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise. Color perception of shade is \ktm{influenced by the contrasting effect of surrounding shades}, allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend \ktm{\citep{bartleson1967brightness,adelson1993perceptual,fairchild2013color}}. See Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors. It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.  %Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.
